---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: Adam Paul
date: 5-6-2022
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(mosaicData) 
```

```{r loading data, echo=TRUE, collapse=TRUE, message=FALSE}
staff <- read_csv("data/instructional-staff.csv")
```

```{r making the data long}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```



### Exercise 1

```{r starting line graph}
staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type)) +
  geom_line() +
  labs(
    title= "Instructional Staff Employment Trends",
    x= "Employment Year",
    y= "% Employed",
    color= "Faculty Type"
  )

```

>Include the line plot you made above in your report and make sure the figure width is large enough to make it legible. Also fix the title, axis labels, and legend label.

>Suppose the objective of this plot was to show that the proportion of part-time faculty have gone up over time compared to other instructional staff types. What changes would you propose making to this plot to tell this story?

I think the graph effectively conveys this information. However, it can also be portrayed by using a stack chart. Colors are picked from a colorblind friendly palette. 

```{r stack}

staff_long %>%
  ggplot(aes(x= value, y=year, fill= faculty_type)) +
           geom_col() +
  scale_fill_manual(values = c( 
    "Full-Time Tenured Faculty" = "#E69F00",
   "Full-Time Tenure-Track Faculty" = "#56B4E9",
   "Full-Time Non-Tenure-Track Faculty" = "#009E73",
    "Part-Time Faculty" = "#D55E00",
    "Graduate Student Employees" = "#0072B2"))

```

### Exercise 2

There is a terrible fisheries graph.

```{r bringing in fisheries, echo=TRUE, collapse=TRUE, message=FALSE}
fisheries <- read_csv("data/fisheries.csv")
```

>Can you help them make improve it? First, brainstorm how you would improve it. Then create the improved visualization and write up the changes/decisions you made as bullet points. It’s ok if some of your improvements are aspirational, i.e. you don’t know how to implement it, but you think it’s a good idea. Ask a tutor for help, but also keep an eye on the time. Implement what you can and leave note identifying the aspirational improvements.

Okay, the first thing I notice is that there's way too much going on. In talking with Elayna and Marq they both suggested that they cut the number of countries to only those above a certain tonnage. However, I tried a few lengths and found that most left too much on the graph. We really only care about the massive producers, so I cut it up to 900,000 tonnes or above.

```{r big fish only}
bigger_fisheries <- fisheries %>%
  filter(total>900000)
```

With that, I want to step towards looking at the graph. The chunk code to make the graph appear nicely is wholesale stolen from them. Thanks Elayna!

```{r fig.align="center", echo = FALSE, fig.width= 25, fig.height= 30}
bigger_fisheries %>%
  ggplot(aes(x= reorder(country, total), # This orders the countries in ascending order
             y= total, fill = capture)) +
  geom_bar(stat ="identity") +
    coord_flip() + # I don't actually like the way it looks this direction but the names overlap too much the other way
  labs(title = "Fish Production by Nation", 
       subtitle = "All Nations larger than 800,000 tonnes",
       x = "Country", y= "Total (in tonnes)") +
  theme( #trbl are the margin signs, top, right, bottom, left
    plot.title = element_text(size = 30, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 25, margin = margin(b= 10)),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 25, hjust = 1),    
    axis.title.x = element_text(size= 25, margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text.y = element_text(margin = margin(r = 5)),
    axis.text.x = element_text(margin = margin(t = 10)),
  ) +
  scale_y_continuous(labels = scales::comma) + scale_fill_continuous(labels = scales::comma)
```

I think this graph shows the rapid shift from even very large producers (>900,000) to the incredible production of China. The labels are much more readable, the titles informative, and the gradient effective redundancy.


### Exercise 3

```{r load data}
data(Whickham)
```


```{r looking at data}
?Whickham
library(performance)
```


>What type of study do you think these data come from: observational or experiment? Why?

Observational, because they are looking at life trajectories without any intervention.


### Exercise 4

>How many observations are in this dataset? What does each observation represent?

Each observation is an individual person, and there 1314 observations, so 1314 participants in the study.

### Exercise 5

> How many variables are in this dataset? What type of variable is each? Display each variable using an appropriate visualization.

The dataset has three variables, age, outcome, smoker. Age is an integer, and the other two are factors.

```{r variable details}

class(Whickham$age)
class(Whickham$outcome)
class(Whickham$smoker)

```


### Exercise 6

>What would you expect the relationship between smoking status and health outcome to be?

I would expect that smoking status is associated with a higher mortality rate.

### Exercise 7

>Create a visualization depicting the relationship between smoking status and health outcome. Briefly describe the relationship, and evaluate whether this meets your expectations. Additionally, calculate the relevant conditional probabilities to help your narrative. Here is some code to get you started:


```{r graph}
Whickham %>% 
  ggplot(aes(x = smoker, fill = outcome)) +
  geom_bar() +
  labs(title = "Health Outcomes of Smokers and Non-Smokers in Whickham" , y = "Number of People")
```

First off, there are more smokers than non-smokers. Secondly, the proportion of smokers still alive is larger than the proportion of non-smokers that are still alive. This does not go in the direction I expected.


```{r}
Whickham %>% 
  count(smoker, outcome) %>% 
  group_by(smoker) %>% 
  mutate(outcome_prob = n / sum(n))
```


>Create a new variable called age_cat using the following scheme:

```{r adding age_cat}

Whickham <- Whickham %>%
  mutate(age_cat=case_when(
age <= 44 ~ "18-44",
age > 44 & age <= 64 ~ "45-64",
age > 64 ~ "65+"))
```

### Exercise 8

>Re-create the visualization depicting the relationship between smoking status and health outcome, faceted by age_cat. What changed? What might explain this change? Extend the contingency table from earlier by breaking it down by age category and use it to help your narrative.

```{r }
Whickham %>% 
  ggplot(aes(x= smoker, fill= outcome)) +
  geom_bar()+
  facet_wrap(~age_cat)
```

```{r age_cat vis}
Whickham %>%
  count(smoker, age_cat, outcome) %>%
    group_by(smoker, age_cat) %>% 
  mutate(percentage = n / sum(n))
```

What we see is that young people tend to be alive, and older people tend to be dead. But notably, there are much, much less older smokers, meaning they collected data from less smokers. But it also means they aren't making it to that age as often. And to support that, they do in fact die more than non-smokers when they're in the 45-65.

